---
title: "Hands-On Tutorials for Covariate Adjustment in Randomized Trials"
subtitle: "Binary Outcome: Standardization Estimator"
author: "Josh Betz (jbetz@jhu.edu), Kelly Van Lancker (kelly.vanlancker@ugent.be), and Michael Rosenblum (mrosen@jhu.edu)"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
header-includes:
   - \usepackage{amsmath, bm}
   - \usepackage{unicode-math}
output:
  ioslides_presentation:
    css: [!expr 'system.file(package = "rmarkdown", "rmd", "h", "bootstrap", "css", "bootstrap.css")', 'styles.css']
    widescreen: true
bibliography:
  - "../bibtex_files/covariate_adjustment.bib"
---

```{r Setup, include = FALSE}
overwrite_results <- FALSE
cached_results_file <- "binary_sim_mistie_iii_standardization.Rdata"
cached_results_path <-
  file.path(
    "C:", "Users", "jbetz", "OneDrive - Johns Hopkins", "Documents",
    "Applied_Covariate_Adjustment_Workshop", cached_results_file
  )

rng_seed <- 17386946

# Quantile Type
#  - 2 = SAS, Stata commands `summarize`, `detail`, `xtile`, `pctile`
#  - 6 = Minitab, SPSS, Stata `centile`
#  - 7 = S, R default
q_type <- 7
stat_digits <- 2
na_is_category <- FALSE # Treat NA as level of variable in tabulations

n_participants <- 500 # Number of participants to use
show_data_rows <- 10 # Number of rows of raw data to show
table_bootstrap_options = c("striped", "hover") # HTML Report Bootstrap options

alpha <- 0.05

# Summary Stats to Compute
compute_mean_sd = TRUE
compute_median_iqr = TRUE
compute_range = FALSE



### Set Default Options ########################################################
options(
  knitr.kable.NA = "",
  width = 300
)




### Set Default Options ########################################################
fig_w <- 8
fig_h <- 8

knitr::opts_chunk$set(
  echo = FALSE,
  results = "markup",
  fig.width = fig_w,
  fig.height = fig_h,
  fig.align = "center",
  out.width = "90%",
  out.height = "90%",
  dpi = 600,
  dev = "CairoPNG"
)

### options ####################################################################
options(
  knitr.kable.NA = ''
)
```




```{r Load Packages}
suppressWarnings(
  expr =
    suppressMessages(
      expr = {
        library(tidyr)
        library(dplyr)
        library(table1)
        library(cobalt)
        library(kableExtra)
        library(ggplot2)
        library(boot)
      }
    )
)
```




```{r Load Cached Results}
results_exist <- file.exists(cached_results_path)
run_analyses <- !results_exist | overwrite_results

if(results_exist){
  load(file = cached_results_path)
}
```




```{r Supplementary Code, eval = run_analyses}
## Note: These are only for producing formatted tables.
render_categorical <-
  function(
    x,
    na.is.category = na_is_category
  ) {
    return(
      render.categorical.default(
        x = x,
        na.is.category = na.is.category
      )
    )
  }


render_continuous <-
  function(
    x,
    mean_sd = compute_mean_sd,
    median_iqr = compute_median_iqr,
    range = compute_range,
    quantile_type = q_type,
    digits = stat_digits,
    na.rm = TRUE
  ) {
    return(
      c("", 
        # Compute/Format Mean (SD)
        if(mean_sd){
          mean_sd <- 
            sprintf(
              fmt = paste0("%.", digits, "f"),
              c(mean(x, na.rm = na.rm), sd(x, na.rm = na.rm)
              )
            )
          
          c("Mean (SD)" =
              base::sprintf(
                fmt = "%s (%s)", mean_sd[1], mean_sd[2]
              )
          )
        } else {
          NULL
        },
        # Compute/Format Median [IQR]
        if(median_iqr){
          median_iqr <- 
            sprintf(
              fmt = paste0("%.", digits, "f"),
              quantile(x = x, p = c(0.50, 0.25, 0.75), na.rm = na.rm)
            )
          
          c("Median [IQR]"=
              base::sprintf(
                fmt = "%s [%s, %s]", median_iqr[1], median_iqr[2], median_iqr[3]
              )
          )
        } else {
          NULL
        },
        
        # Compute/Format Range
        if(range){
          range <- 
            sprintf(
              fmt = paste0("%.", digits, "f"),
              base::range(x = x, na.rm = na.rm)
            )
          c("Range" = 
              base::sprintf(
                fmt = "[%s, %s]", range[1], range[2]
              )
          )
        } else{
          NULL
        }
      )
    )
  }


ate_binary <-
  function(
    data,
    indices = NULL,
    formula,
    link = c("logit", "probit", "cauchit")[1],
    tx_var = "tx"
  ){
    # When bootstrap indices are supplied, 
    if(!is.null(indices)) data <- data[indices,]
    
    # Check if `tx_var` is in `formula`:
    if(!(tx_var %in% labels(terms(formula)))) {
      stop(tx_var, " is not in the formula: ", base::deparse1(expr = formula))
    }
    
    # Fit the model
    outcome_model <-
      stats::glm(
        formula = formula,
        family = stats::binomial(link = link),
        data = data
      )
    
    # Predict each individual's outcome under control
    mu_0 <-
      pr_outcome_unadj_control <-
      stats::predict(
        object = outcome_model,
        newdata = 
          base::within(
            data = data,
            expr = {
              assign(x = tx_var, value = 0)
            }
          ),
        type = "response"
      )
    
    # Predict each individual's outcome under treatment
    mu_1 <-
      pr_outcome_unadj_control <-
      stats::predict(
        object = outcome_model,
        newdata = 
          base::within(
            data = data,
            expr = {
              assign(x = tx_var, value = 1)
            }
          ),
        type = "response"
      )
    
    # Average predictions over sample
    e_y_0 <- mean(mu_0)
    e_y_1 <- mean(mu_1)
    
    # Produce contrasts
    return(
      c("RD" = e_y_1 - e_y_0, "RR" = e_y_1/e_y_0,
        "OR" = (e_y_1*(1 - e_y_0))/(e_y_0*(1 - e_y_1)),
        "E[Y|A=1]" = e_y_1, "E[Y|A=0]" = e_y_0)
    )
  }


all_boot_cis <-
  function(boot_object, alpha = 0.05, method = "bca"){
    all_cis <- list()
    n_params <- length(boot_object$t0)
    
    for(i in 1:n_params){
      all_cis[[i]] <- 
        boot::boot.ci(
          boot.out = boot_object,
          conf = 1 - alpha,
          type = method,
          index = i
        )
    }
    
    all_cis <-
      sapply(
        X = all_cis,
        FUN = function(ci) tail(get(x = method, pos = ci)[1,], 2)
      )
    
    return(
      data.frame(
        "Estimate" = boot_object$t0,
        "SE" = base::apply(X = boot_object$t, MARGIN = 2, FUN = sd),
        "Var" = base::apply(X = boot_object$t, MARGIN = 2, FUN = var),
        "LCL" = all_cis[1,],
        "UCL" = all_cis[2,],
        "CI Width" = all_cis[2,] - all_cis[1,],
        check.names = FALSE
      )
    )
  }
```


```{r Process Data, eval = run_analyses}
data_url <-
  paste0("https://github.com/jbetz-jhu/CovariateAdjustmentTutorial",
         "/raw/main/Simulated_MISTIE_III_v1.2.csv")

sim_miii_full <- read.csv(file = url(data_url))

sim_miii_full <-
  sim_miii_full %>% 
  dplyr::tibble() %>% 
  dplyr::mutate(
    # Convert variables from binary indicators to labeled categorical variables
    male =
      factor(
        x = male,
        levels = 0:1,
        labels = c("0. Female", "1. Male")
      ),
    
    across(
      .cols = 
        all_of(
          x = c("hx_cvd", "hx_hyperlipidemia",
                "on_anticoagulants", "on_antiplatelets")
        ),
      .fns = function(x) factor(x, levels = 0:1, labels = c("0. No", "1. Yes"))
    ),
    
    # Convert GCS and MRS variables from character data to categorical variables
    across(
      .cols = starts_with("gcs") | starts_with("mrs"),
      .fns = factor
    ),
    
    ich_location =
      factor(
        x = ich_location,
        levels = c("Deep", "Lobar")
      ),
    
    arm =
      factor(
        x = arm,
        levels = c("medical", "surgical")
      ),
    
    tx = 1*(arm == "surgical"),
    
    mrs_356d_binary =
      case_when(
        mrs_365d %in% c("0-1", "2", "3") ~ 1,
        mrs_365d %in% c("4", "5", "6") ~ 0,
      ),
    
    mrs_356d_binary_factor =
      factor(
        x = mrs_356d_binary
      )
  )


# Take the first 500 rows
sim_miii <-
  sim_miii_full %>% 
  dplyr::slice(1:n_participants)

# Apply labels to Variables
label(sim_miii$age) <- "Age at Presentation (y)"
label(sim_miii$male) <- "Sex"
label(sim_miii$hx_cvd) <- "Hx Cardiovascular Disease"
label(sim_miii$hx_hyperlipidemia) <- "Hx Hyperlipidemia"
label(sim_miii$on_anticoagulants) <- "On Anticoagulants: Presentation"
label(sim_miii$on_antiplatelets) <- "On Antiplatelets: Presentation"

label(sim_miii$ich_location) <- "ICH Location"
label(sim_miii$ich_s_volume) <- "ICH Volume (mL): Stability"
label(sim_miii$ivh_s_volume) <- "IVH Volume (mL): Stability"
label(sim_miii$gcs_category) <- "Glasgow Coma Scale: Presentation"

label(sim_miii$mrs_30d) <- "Modified Rankin Scale: 30 Days"
label(sim_miii$mrs_180d) <- "Modified Rankin Scale: 180 Days"
label(sim_miii$mrs_365d) <- "Modified Rankin Scale: 365 Days"
label(sim_miii$mrs_356d_binary) <- "Modified Rankin Scale < 4: 365 Days"
label(sim_miii$mrs_356d_binary_factor) <- "Modified Rankin Scale < 4: 365 Days"
```


```{r Unadjusted Analysis, eval = run_analyses}
set.seed(seed = rng_seed)

mrs_unadjusted_logistic_glm <-
  stats::glm(
    formula = 
      mrs_356d_binary ~ arm,
    data = sim_miii,
    family = binomial(link = "logit")
  )

ate_unadj_boot <-
  boot::boot(
    data = sim_miii,
    statistic = ate_binary,
    # Number of Bootstrap Replicates
    R = 10000,
    formula = mrs_356d_binary ~ tx,
    link = "logit",
    tx_var = "tx"
  )

ate_unadj_results <-
  all_boot_cis(ate_unadj_boot)
```


```{r Adjusted Analysis, eval = run_analyses}
mrs_adjusted_logistic_glm <-
  stats::glm(
    formula = 
      mrs_356d_binary ~ arm +
      age + male + hx_cvd + hx_hyperlipidemia +
      on_anticoagulants + on_antiplatelets +
      ich_location + ich_s_volume + ivh_s_volume + 
      gcs_category,
    data = sim_miii,
    family = binomial(link = "logit")
  )

ate_adj_boot <-
  boot::boot(
    data = sim_miii,
    statistic = ate_binary,
    R = 10000,
    formula = 
      mrs_356d_binary ~ tx +
      age + male + hx_cvd + hx_hyperlipidemia +
      on_anticoagulants + on_antiplatelets +
      ich_location + ich_s_volume + ivh_s_volume + 
      gcs_category,
    link = "logit",
    tx_var = "tx"
  )

ate_adj_results <-
  all_boot_cis(ate_adj_boot)
```

```{r Compare Unadjusted to Adjusted}
# Variance = SE^2
var_adjusted <- 
  base::apply(X = ate_adj_boot$t, MARGIN = 2, FUN = var)
var_unadjusted <- 
  base::apply(X = ate_unadj_boot$t, MARGIN = 2, FUN = var)

var_adjusted <-
  setNames(
    object = var_adjusted,
    nm = c("RD", "RR", "OR", "E[Y|A=1]", "E[Y|A=0]")
  )

var_unadjusted <-
  setNames(
    object = var_unadjusted,
    nm = c("RD", "RR", "OR", "E[Y|A=1]", "E[Y|A=0]")
  )

# Adjusted asymptotically at least as efficient as unadjusted:
relative_efficency_adj_unadj <-
  var_unadjusted/var_adjusted

relative_change_precision <-
  relative_efficency_adj_unadj - 1

percent_change_precision <-
  100*relative_change_precision

relative_change_variance <-
  (1/relative_efficency_adj_unadj) - 1

percent_change_variance <-
  100*relative_change_variance

relative_ci_width <-
  ate_adj_results$`CI Width`/ate_unadj_results$`CI Width`

covariate_adjustment_changes <-
  dplyr::tibble(
    "Estimate" = names(var_adjusted),
    "Relative Efficiency" = relative_efficency_adj_unadj,
    "Relative Change in Precision" = relative_change_precision,
    "Percent Change in Precision" = percent_change_precision,
    "Relative Change in Variance" = relative_change_variance,
    "Percent Change in Variance" = percent_change_variance,
    "Relative CI width" = relative_ci_width
  )
```

```{r Cache Results, eval = run_analyses}
save(
  file = cached_results_path,
  list =
    c(
      "sim_miii_full",
      "sim_miii",
      "render_categorical",
      "render_continuous",
      "ate_binary",
      "all_boot_cis",
      "mrs_unadjusted_logistic_glm",
      "ate_unadj_boot",
      "ate_unadj_results",
      "mrs_adjusted_logistic_glm",
      "ate_adj_boot",
      "ate_adj_results",
      "var_adjusted",
      "var_unadjusted",
      "covariate_adjustment_changes"
    )
)
```


<style>
.forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>




## Outline:

  - Covariate Adjustment Tutorials page
  - Worked Example: Binary Outcome - Standardization Estimator
    - Example Data: MISTIE III trial - Hemorrhagic Stroke
    - Outcome: Modified Rankin Score - Dichotomized
    - Unadjusted & Covariate Adjusted Analyses
    - Confidence Intervals: Nonparametric Bootstrap


# Hands-On Tutorials for Covariate Adjustment | [https://bit.ly/rct_tutorials](https://bit.ly/rct_tutorials)




## Tutorials on Covariate Adjustment | [https://bit.ly/rct_tutorials](https://bit.ly/rct_tutorials)

  - Simulated data: mimic features of trials - Scale, correlations, missingness
    - Continuous, Ordinal, Binary, Time-to-Event outcomes: Covariates
    - Example datasets: Substance Abuse, Hemorrhagic Stroke
    - Stratified randomization: improve precision
  - Common estimands of interest; Analytic approaches
  - R code for tabulating, plotting, analyzing data:
    - Unadjusted & Adjusted
  - Links to resources on learning and using R




# Example: based on MISTIE III | Functional Outcome & Mortality in Hemorrhagic Stroke




## MISTIE-III Trial: [@Hanley2019]

  - Hemorrhagic Stroke: Greater morbidity, mortality than ischemic stroke
    - Intracerebral Hemorrhage (ICH), possibly with Intraventricular Hemorrhage (IVH)
  - Consent: monitor daily for ICH stability by CT
  - 1:1 randomized - minimally invasive surgery + thrombolytic vs. SOC medical management
  - Safety & Efficacy: Functional outcome on Modified Rankin Scale (MRS)
    - MRS at 30, 180, and 365 days post randomization
  - Good Outcome: MRS 0-3 vs. 4-6 - independent vs. not
  - Simulated data based on actual trial data: **not actual study data.**




## Simulated MISTIE Data: {.columns-2 .smaller}

  - [Codebook on covariateadjustment.github.io](https://covariateadjustment.github.io/index.html#mistie_iii)
  - Baseline Covariates
    - `age`: Age in years
    - `male`: male sex
    - `hx_cvd`:	cardiovascular disease history
    - `hx_hyperlipidemia`:	hyperlipidemia
    - `on_anticoagulants`:	on anticoagulants
    - `on_antiplatelets`:	on antiplatelets
    - `ich_location`: ICH: (`Lobar`, `Deep`)
    - `ich_s_volume`:	ICH volume on stability scan
    - `ivh_s_volume`:	IVH volume on stability scan
    - `gcs_category`: presenting Glasgow Coma Score


<p class="forceBreak"></p>


  - Treatment:
    - `arm`: treatment arm
    - `ich_eot_volume`: intracerebral hemorrhage volume on end-of-treatment scan
  - Outcome:
    - Modified Rankin: `_complete`: completely observed
    - `mrs_30d`: MRS at 30 days (`0-3`, `4`, `5`, `6`)
    - `mrs_180d`: MRS at 180 days (`0-2`, `3`, `4`, `5`, `6`)
    - **`mrs_365d`: MRS at 365 days (`0-1`, `2`, `3`, `4`, `5`, `6`)**
      - **Primary Outcome**
    - `days_on_study`: days until death/censoring
    - `died_on_study`: participant died (`1`) or censored (`0`)




## Standardization Estimator {.smaller}

  - $Y$ denotes the outcome: 1 = 1-Year MRS 0-3; 0 = 1-Year MRS 4-6
  - $A$ denotes treatment assignment: 1 = Treatment, 0 = Control
  - **Fit** a regression model for the outcome:
    - Undjusted: $logit(Pr\{Y = 1 \vert A \}) = \beta_{0} + \beta_{A}A$
    - Adjusted: $logit(Pr\{Y = 1 \vert A \}) = \beta_{0} + \beta_{A}A + \beta_{1}X_{1} + \ldots \beta_{p}X_{p}$
  - **Predict** each individual's outcome using the fitted model
    - $\hat{y}^{(1)}_{i} = logit^{-1}\{\hat{\beta}_{0} + \hat{\beta}_{A} + \hat{\beta}_{1}X_{i1} + \ldots \hat{\beta}_{p}X_{ip}\}$
    - $\hat{y}^{(0)}_{i} = logit^{-1}\{\hat{\beta}_{0} + \hat{\beta}_{1}X_{i1} + \ldots \hat{\beta}_{p}X_{ip}\}$
  - **Average** these predictions over the sample
    - $\hat{\mu}^{(1)} = \frac{1}{n}\sum_{i=1}^{n}\hat{y}^{(1)}_{i} \qquad \hat{\mu}^{(0)} = \frac{1}{n}\sum_{i=1}^{n}\hat{y}^{(0)}_{i}$
  - **Contrast** the averaged predictions
    - $\hat{\theta}_{RD} = \hat{\mu}^{(1)} - \hat{\mu}^{(0)} \qquad \hat{\theta}_{RR} = \hat{\mu}^{(1)}/\hat{\mu}^{(0)}$




# Unadjusted Analysis | Standardization (also known as G-Computation)




## Fitting Logistic Model {.smaller}

  - [See tutorials for more on GLMs in R](https://covariateadjustment.github.io/using_r.html#fitting-regression-models)
```{r Fit Unadjusted Model, echo = TRUE}
mrs_unadjusted_logistic_glm <-
  stats::glm(
    formula = 
      mrs_356d_binary ~ arm,
    data = sim_miii,
    family = binomial(link = "logit")
  )

summary(mrs_unadjusted_logistic_glm)
```




## Predict Outcomes {.columns-2 .smaller}

```{r Generate Unadjusted Predictions, echo = TRUE}
pr_outcome_unadj_control <-
  stats::predict(
    object = mrs_unadjusted_logistic_glm,
    newdata = 
      within(data = sim_miii,
             expr = {arm = "medical"}),
    type = "response"
  )

pr_outcome_unadj_treatment <-
  stats::predict(
    object = mrs_unadjusted_logistic_glm,
    newdata = 
      within(data = sim_miii,
             expr = {arm = "surgical"}),
    type = "response"
  )
```


<p class="forceBreak"></p>


```{r Show Unadjusted Predictions, echo = TRUE}
table(pr_outcome_unadj_control)
table(pr_outcome_unadj_treatment)
```

  - Treatment is only covariate in model
  - Predictions are generated for each person assigning them to each treatment
  - Unadjusted predictions will be identical




## Average & Contrast Predictions  {.columns-2 .smaller}

```{r Average & Contrast Unadjusted Predictions 1, echo = TRUE}
e_y_0_unadj <- mean(pr_outcome_unadj_control)
e_y_1_unadj <- mean(pr_outcome_unadj_treatment)

# Risk Difference
e_y_1_unadj - e_y_0_unadj
# Relative Risk
e_y_1_unadj/e_y_0_unadj
# Odds Ratio
(e_y_1_unadj*(1 - e_y_0_unadj))/
  (e_y_0_unadj*(1 - e_y_1_unadj))
```


<p class="forceBreak"></p>

  - Compare two counterfactual worlds using information from each arm:
    - All eligible patients in the population receive surgical intervention
    - All eligible patients in the population receive medical intervention
    - Probability of good outcome is 3% higher in population if everyone receives surgical intervention than if everyone received standard medical care.
    - Probability of "good outcome" is 7% greater: Ratio
    - Odds of a "good outcome" is 13% higher: Ratio
  - Odds Ratio overestimates relative risk: Outcome is not rare



## Average & Contrast Predictions {.columns-2 .smaller}

```{r Average & Contrast Unadjusted Predictions 2, echo = TRUE}
e_y_0_unadj <- mean(pr_outcome_unadj_control)
e_y_1_unadj <- mean(pr_outcome_unadj_treatment)

# Risk Difference
e_y_1_unadj - e_y_0_unadj
# Relative Risk
e_y_1_unadj/e_y_0_unadj
# Odds Ratio
(e_y_1_unadj*(1 - e_y_0_unadj))/
  (e_y_0_unadj*(1 - e_y_1_unadj))
```


<p class="forceBreak"></p>


```{r Compute Unadjusted Contrasts from Model, echo = TRUE}
unadj_glm_beta <- coef(mrs_unadjusted_logistic_glm)
pr_medical <-
  plogis(unadj_glm_beta["(Intercept)"])

pr_surgical <-
  plogis(unadj_glm_beta["(Intercept)"] +
           unadj_glm_beta["armsurgical"])

pr_surgical - pr_medical # Risk Difference
pr_surgical/pr_medical # Relative Risk
exp(unadj_glm_beta["armsurgical"]) # Odds Ratio
```




## Compute CIs using Bootstrap

<div style="float: left; width: 40%;">
<small>
  - For `ate_binary()` code, see the workshop materials

```{r Compute Unadjusted ATE, eval = FALSE, echo = TRUE}
ate_unadj_boot <-
  boot::boot(
    data = sim_miii,
    statistic = ate_binary,
    R = 10000,
    formula =
      mrs_356d_binary ~ tx,
    link = "logit",
    tx_var = "tx"
  )

ate_unadj_results <-
  all_boot_cis(ate_unadj_boot)
```

</small>
</div>


<div style="float: left; width: 60%;">

```{r Show Unadjusted ATE, echo = FALSE, out.width="100%"}
ate_unadj_results %>% 
  kableExtra::kbl(
    caption = 
      "Unadjusted Estimates: Standardization",
    digits = c(2, 2, 4, 2, 2, 2)
  ) %>% 
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = table_bootstrap_options
  )
```

</div>




# Covariate adjusted Analysis | Standardization (also known as G-Computation)




## Unadjusted vs Adjusted Standardization Estimates

  - Standardization Estimates vs. Logistic Regression Output
    - Same estimates of RD, RR, and OR
  - This will not be true in general for adjusted analyses:
    - Regression: **conditional** - Standardization: **marginal**
  - Only change necessary for adjusted analysis: Add covariates to model
    - All other steps identical




## Adding Covariates to Logistic Model {.smaller}

<div style="float: left; width: 35%;">

```{r Fit Adjusted Model 1, echo = TRUE, eval = FALSE, out.width = "30%"}
mrs_adjusted_logistic_glm <-
  stats::glm(
    formula = 
      mrs_356d_binary ~ arm +
      age + 
      male +
      hx_cvd +
      hx_hyperlipidemia +
      on_anticoagulants +
      on_antiplatelets +
      ich_location +
      ich_s_volume +
      ivh_s_volume + 
      gcs_category,
    data = sim_miii,
    family =
      binomial(link = "logit")
  )
```

</div>


<div style="float: right; width: 65%;">

```{r Fit Adjusted Model 2, echo = FALSE, eval = TRUE, out.width = "70%"}
mrs_adjusted_logistic_glm <-
  stats::glm(
    formula = 
      mrs_356d_binary ~ arm +
      age + 
      male +
      hx_cvd +
      hx_hyperlipidemia +
      on_anticoagulants +
      on_antiplatelets +
      ich_location +
      ich_s_volume +
      ivh_s_volume + 
      gcs_category,
    data = sim_miii,
    family = binomial(link = "logit")
  )

summary(mrs_adjusted_logistic_glm)
```

</div>


## Covariate Adjusted Estimate: Standardization

<div style="float: left; width: 40%;">

```{r Compute Adjusted ATE, eval = FALSE, echo = TRUE}
ate_adj_boot <-
  boot::boot(
    data = sim_miii,
    statistic = ate_binary,
    # Number of Bootstrap Replicates
    R = 10000,
    formula = 
      mrs_356d_binary ~ tx +
      age + male + hx_cvd +
      hx_hyperlipidemia +
      on_anticoagulants +
      on_antiplatelets +
      ich_location +
      ich_s_volume +
      ivh_s_volume + 
      gcs_category,
    link = "logit",
    tx_var = "tx"
  )

ate_adj_results <-
  all_boot_cis(ate_adj_boot)
```

</div>


<div style="float: left; width: 60%;">

<div style="float: left; width: 60%;">

```{r Show Adjusted ATE, echo = FALSE, out.width="100%"}
ate_adj_results %>% 
  kableExtra::kbl(
    caption = 
      "Adjusted Estimates: Standardization",
    digits = c(2, 2, 4, 2, 2, 2)
  ) %>% 
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = table_bootstrap_options
  )
```

</div>


## Compare Results: Adjusted vs. Unadjusted

```{r Compare Adjusted to Unadjusted}
covariate_adjustment_changes %>% 
  dplyr::select(
    -dplyr::any_of(
      x = c("Percent Change in Precision", "Percent Change in Variance")
    )
  ) %>% 
  kableExtra::kbl(
    caption = "Changes in precision and variance from covariate adjustment.",
    digits = c(0, 2, 3, 1, 2, 1, 2)
  ) %>% 
  kableExtra::kable_styling(
    bootstrap_options = table_bootstrap_options
  )
```


## Summary {.smaller}

  - Standardization always gives a marginal treatment effect:
    - Same estimand/target with or without covariate adjustment
  - Logistic regression 
    - No covariates: **marginal** treatment effect
    - Include covariates: **conditional** treatment effect
  - Conditional & Marginal rarely coincide except with a null effect or linear link
  - Bootstrap CI gives appropriate coverage: Important for testing
    - Does not assume logistic model is correctly specified
    - Adjusted analysis at least as efficient as unadjusted analysis asymptotically




## References {.smaller}